Dynamics of Laterally Propagating Flames in X-ray Bursts. III. Comparing Early Hotspot Evolution in Two and Three Dimensions

### 2023-02-27 xrb3/paper.tex
more work

X-ray bursts (XRBs) result from thermonuclear burning of an accreted
H/He or He layer on a neutron star \citep{galloway:2017}.
Observations of brightness oscillations in the rise of the lightcurve
suggest that the burning begins localized and then spreads across the
neutron star.  This spreading is inherently a multi-dimensional
phenomena, and hydrodynamic simulations that resolve the reactive
zone, use realistic reaction networks, and capture the scales over
which rotation is important are needed to understand the flame
propagation and nucleosynthesis.  We show a first attempt at modeling
the early evolution of a spreading hotspot in three dimensions.  This
builds on our earlier two dimensional work
\citep{eiden:2020,harpole:2021} that developed our simulation
framework and explored how the acceleration of the burning front
depends on the initial model structure.  As with those simulations, we
use the freely-available \castro\ simulation code
\citep{castro,castro_joss} simulation code, with all of the code
needed to run these simulations in our public github repos.

@@

Figure~\ref{fig:vr_abar} shows the a three-dimensional simulation via
top-down volume rendering of the mean-molecular weight,
\begin{equation}
\bar{A} = \sum_k \frac{X_k}{A_k}
\end{equation}
where $X_k$ is the mass fraction of species $k$ and $A_k$ is the
atomic weight (in mass units).  The structure is shown at 10, 20, and
40~ms of simulation time.  While initially quite symmetric, the axial
symmetry is broken and the flame structure takes on a more tenuous
form.  The panels show the flame spreading considerably through the
domain as time evolves, and the simulation is halted once the flame
nears the boundaries.

@@

There is still a role for three-dimensional simulations---the
convective burning in the accreted layer should create turbulence that
the flame will encounter as it propagates.  Although we have shown
that for the flame in a Type Ia supernovae, the turbulence from the
era of convection is not strong enough to disrupt the
flame~\cite{wdturb}, the flame in an XRB is considerably slower and
thicker, so it remains to be seen what effect it might have.  This can
be assessed in a different geometry, like a long channel, which would
allow us to trade domain size for resolution to increase the numerical
Reynolds number.  We also will consider higher-order simulation
methodologies which could allow us to capture these effects at lower
resolution.  We've already demonstrated a fully fourth-order accruate
(in space and time) algorithm for coupling hydro, diffusion, and
reactions in Castro \cite{castro-sdc} that could be used here.  The
main outstanding issue with applying that work to the present problem
is the multilevel time integration.

>>>>

X-ray bursts (XRBs) result from thermonuclear burning of an accreted
H/He or He layer on a neutron star \citep{galloway:2017}.
Observations of brightness oscillations in the rise of the lightcurve
suggest that the burning begins localized and then spreads across the
neutron star \citep{bhattacharyya:2007}.  This spreading is inherently a multi-dimensional
phenomena, and hydrodynamic simulations that resolve the reactive
zone, use realistic reaction networks, and capture the scales over
which rotation is important are needed to understand the flame
propagation and nucleosynthesis (especially rp-process in mixed H/He bursts \citealt{rpprocess}).  We show a first attempt at modeling
the early evolution of a spreading hotspot in three dimensions.  This
builds on our earlier two dimensional work
\citep{eiden:2020,harpole:2021} that developed our simulation
framework and explored how the acceleration of the burning front
depends on the initial model structure.  As with those simulations, we
use the freely-available \castro\ simulation code
\citep{castro,castro_joss} simulation code, with all of the code
needed to run these simulations in our public github repos.

@@

Figure~\ref{fig:vr_abar} shows the a three-dimensional simulation via
top-down volume rendering of the mean-molecular weight,
\begin{equation}
\frac{1}{\bar{A}} = \sum_k \frac{X_k}{A_k}
\end{equation}
where $X_k$ is the mass fraction of species $k$ and $A_k$ is the
atomic weight (in mass units).  The structure is shown at 10, 20, and
40~ms of simulation time.  While initially quite symmetric, the axial
symmetry is broken and the flame structure takes on a more tenuous
form.  The panels show the flame spreading considerably through the
domain as time evolves, and the simulation is halted once the flame
nears the boundaries.

@@

There is still a role for three-dimensional simulations---the
convective burning in the accreted layer should create turbulence that
the flame will encounter as it propagates.  Although we have shown
that for the flame in a Type Ia supernovae, the turbulence from the
era of convection is not strong enough to disrupt the
flame~\citep{wdturb}, the flame in an XRB is considerably slower and
thicker, so it remains to be seen what effect it might have.  This can
be assessed in a different geometry, like a long channel, which would
allow us to trade domain size for resolution to increase the numerical
Reynolds number.  We also will consider higher-order simulation
methodologies which could allow us to capture these effects at lower
resolution.  We've already demonstrated a fully fourth-order accruate
(in space and time) algorithm for coupling hydro, diffusion, and
reactions in Castro \citep{castro-sdc} that could be used here.  The
main outstanding issue with applying that work to the present problem
is the multilevel time integration.
